Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2019-08-12 Response_to_Reviewers.tex

   $\rightarrow$ To describe in more details the density grid generator we added a few sentences; and the generator was furthermore better situated among the many possible others in the discussion.
   %ways of making a spatial generator as Young2002 (FLN/MLT).
 	% item \comment[JR]{will add some description on urban form indicators etc.}

  \medskip

   \item \textit{This also applies to the three segregation indexes: what exactly do they measure?}

  $\rightarrow$ Segregation indexes were better described in the section describing the models.

@@

  \medskip

  \item \textit{The description of the spatial generator’s procedure, outlined in paragraph 2.7, would also benefit if more information is added} \ldots
  %, for example, to delineate the terms “strength of attraction” (attraction in which respect?) and “strength of the diffusion process” (is strength equal to intensity? If so, is it then in space and/or time?).

  $\rightarrow$ The description of the spatial generator was partly reformulated with more details.

@@

  $\rightarrow$ see comment A. cite Banos, 2012 \comment[FL]{quel était le point exactement avec Banos?}

  \medskip

  \textit{The problem of setting space and geography as synonymous continues in paragraphs 1.9 and following. What is described here as "spatial system shape" is the geometry of spaces but not the geography of spaces.}
  % -> yes indeed

  $\rightarrow$ We agree and corrected the semantics (see comment A).

@@

  $\rightarrow$ We agree with this comment and recall that urban growth processes are not the main focus of the paper, they are here an instantiation of a given $\gamma$ generator, which could have been replaced by others such as network models, procedural modeling, etc. The position and discussion of the paper was amended to precise that (see comment A).

  %it's not the point of the paper, only the metaphore of the spatial generator. + supprimer la ref à "urban growth" dans le texte et dire d'inspiration urban science mais dont seules les grilles finales nous intéressent. On aurait pu utiliser des generator issus de graphes, de percolation etc.
  %The core of the paper is not this, it's the approach of including gamma and applying to mu to understand systems better, whatever gamma and whatever mu in whatever knowledge domain.
  %\item in discussion, mention that it's one isolated example of geometry but other cases could be thought of as well, using networks and graphs, but also empirical shape files.

  \medskip

  \item \textit{The approach is reductionist and has its strength in reflecting growth at an abstract level nevertheless. The concluding remark in paragraph 5.1 “With the Schelling experiment, we found … that polycentric and discontinuous cities appear systematically more segregated than compact cities …” is elusive. Either it is a trivial fact, simply because polycentric cities have more centres and thus, by nature, more opportunities from which segregation can originate. Or, the results can be used as a starting point to dig deeper in geographical reasons, taking social, economic, cultural, and political, but also topographical issues into consideration. This geographical reasoning, however, is outside of the scope of this paper. }
  %\comment[FL]{j'ai pas compris pourquoi la phrase du reviewer avait été mise en commentaire, j'ai remis}
 %\comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate} \comment[MLT]{I kind of disagree with the idea that it is straightforward that centralised cities produce more segregation as it is not a direct parameter of the model but an output. Plus: what argument would he or she have to state that a big center would be automatically less segregated that several smaller ones? And I believe that explaining why polycentric cities are more segregated than others is out of the scope of the paper. But we could add his or her comment in the discussion section, saying that we hove we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?} \comment[FK]{du coup je fais une proposition de réponse ci-dessous en reprenant ces différents éléments}
  We understand this statement, however we wish to underline that our objective in the discussion of the Schelling results was not to establish a general result on the links between polycentricity and segregation, which are indeed very complex and multidimensional as advocated by the reviewer. Instead, we aimed at showing what kind of results could be obtained using the framework we suggest. If we do not prove that in general, and in empirical context, polycentricity creates more segregation, we obtained, in the modelling context, under Schelling assumptions, that polycentric geometric structures are more often that not associated with segregated patterns, a result which we do not believe to be solely resulting from the fact that more centers leads to more opportunities of segregation (which is true), because it seems also true that monocentric structures, with higher densities, can more easily accommodate heterogeneity of agents within cells, and hence a more patchy type of segregation which can for instance being observed in London. We included a comment in the discussion section, saying that we hope we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results? \comment[FL]{à la relecture de l'article s'assurer que cette prise de recul est bien explicitée}

  \comment[FL]{ne pas oublier les trois points ci-dessous!!!}
  \item in discussion, say that indeed the indicators can themselves be influenced by geometric structures (cf. comment B) but on the processes, there is nothing obvious about the influence of structures on behavioural spatial patterns (if you imagine a model with no interactions between agents for example) + it's a difference between the model and its potential addition to the real systems. cf. B

@@

\medskip

\begin{enumerate}

  \item \textit{While the different outcomes of the two implementation exercises are not the main point, I would suggest authors to provide some more elaborated explanations. The two models have different behavioural assumptions and represent a different 'interaction' context. How can be extrapolate and isolate spatial effects when models have different logics and theoretical purposes?}

  $\rightarrow$ See above comment A, discussion has been amended and details added.
  \medskip

@@

  $\rightarrow$ See comment B above.

  \medskip

  \textit{For instance, it would be useful to have suggestions on how to use this methodology to run empirical models. Suppose to have a nice dataset and a case of a urban context to be modelled. Could the exploration of different spatial configurations even be used to run 'counter-factual' or 'what if' tests on actual spatial configurations?}

  $\rightarrow$ This point was not explicit enough and we revised the presentation of our approach (which is generic and not empirical) in introduction (especially fig. 1), as already stated in comment B.3 above. We thank reviewer 2 for highlighting this limitation.
  % \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model}

@@

  \medskip

  \item \textit{Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), which could be interesting to see implemented in ABM.}

  $\rightarrow$ We added a paragraph on the limits of our density grid approach in discussion, in which we mention the possibility to use egohoods to better measure segregation.

@@

  \medskip

   \item \textit{Spatial econometrics based on empirical dataset is developing fast. It would be useful to understand how these models can enrich ABM and vice-versa.}

     $\rightarrow$ This points suggest that our work can inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010. We added a paragraph on this in the discussion discussion, and suggested how our approach allows to go back to questioning empirical cases from the model and questioning our concepts ('neighbourhood').

@@

  \medskip

  \item\textit{Furthermore, even among limitations, perhaps authors could discuss data limitations and the fact that social processes (networks, mobility) cannot conflate with spatial dynamics. Here, it is a bit misleading to call "generator parameters" ONLY the spatial ones!}

  $\rightarrow$ We have clarified this issue in introduction but referring to different terms and figure 1 which differentiates between the $\mu$ model and $\gamma$ parameters (see comment A).

@@

 \medskip

  \item\textit{In this respect, I would encourage authors to further develop their closing section, which now does not stimulate readers to imagine implications of authors' findings and message to their own research.} \ldots
  %I would expand this section to provide a more detailed discussion on implications for model makers, experts on social geography and urban scholars who work with ABM, non-specialist in geography who would like to base their models in more proper geographical constraints. And perhaps, authors could suggest future line of research that could show-case their argument.

  $\rightarrow$ The discussion and conclusion was amended in this direction (see comment B).

### 2019-08-12 paper.tex
.

The spatial generator applies an urban morphogenesis model \citep{batty2007cities} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}. An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily into our workflow. Grids are generated through an iterative process which, starting from an empty grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to a patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (i.e. its 8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $t_f$ is reached. The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four generator parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment. We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$. Typical value ranges for the  parameters will be taken as, following \citet{raimbault2018calibration}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$. We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configurations that can be generated.

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
	%\includegraphics[width=\textwidth]{figuresraw/spatialGen.png}
	\includegraphics[width=\textwidth]{figures/Fig2.png}
 \caption{\textbf{Four examples of grids produced by the spatial generator.} The lighter the red, the denser the area. Changing the growth rate $N$ allows to have more or less chaotic shapes (two first compared to the two last grids for example) corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$. Parameter values used here are for the first grid $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 971)$, the second $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 176)$, the third $(\alpha = 1.4,\beta=0.045, N = 102, n_d = 2, t_f = 618)$, and the fourth $(\alpha = 1.8,\beta=0.114, N = 108, n_d = 1, t_f = 227)$.}
\label{fig:spatialGen}
\end{figure} %
%%%%%%%%%%%%%%%%%%%%%%

% Generator parameters values:
%  1) ex_sp-diffusion=0.0060_sp-growth-rate=25_sp-diffusion-steps=1_sp-alpha-localization=0.4_ticks=971_sp-population=24275.000000000102
%  2) ex_sp-diffusion=0.0060_sp-growth-rate=25_sp-diffusion-steps=1_sp-alpha-localization=0.4_ticks=176_sp-population=4400.000000000003
%  3) ex_sp-max-pop=80260_sp-diffusion=0.045_sp-growth-rate=102_sp-diffusion-steps=2_sp-alpha-localization=1.4_ticks=618
%  4) ex_sp-max-pop=382170_sp-diffusion=0.114_sp-growth-rate=108_sp-diffusion-steps=1_sp-alpha-localization=1.8_ticks=227

In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure: their concentration index, hierarchy index, centrality index and continuity index (cf. \cite{LeNechet2015}). We sample the generator parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a scatter with high discrepancy. We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}. It yields a subset of 170 grids matching empirical densities, which constituted our set of different initial spatial conditions. These are further clustered into three classes of morphology (figure \ref{fig:densityTypes}): compact (e.g. Vienna), polycentric (Liege) and discontinuous (Augsburg). This clustering allows to evaluate the non-trivial effects of a meaningful urban form on simulation results. We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model (cf. section \ref{sec:qualResults}).

>>>>

The spatial generator applies an urban morphogenesis model \citep{batty2007cities} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}. An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily into our workflow. Population density grids, at the typical scale of a metropolitan area, are generated by combining the opposite processes of urban dispersal (negative externalities) with urban concentration (positive externalities). More precisely, grids are generated through an iterative process which simulates successive time steps with a fixed population gain at each time step. Starting from an empty grid, the model adds a quantity $N$ (population) at each time step $t$. The new population is allocated through preferential attachment on previous population density. Formally, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to a patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. The attachment parameter $\alpha$ can thus be interpreted as a ``strength of attraction'', in the sense that increasing it will lead to a higher instantaneous concentration. At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (i.e. its 8 surrounding patches). The parameter $\beta$ can be interpreted as a strength of diffusion: increasing it will lead to larger share of population being diffused in space. To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $t_f$ is reached. The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four generator parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment. We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$. Typical value ranges for the  parameters will be taken as, following \citet{raimbault2018calibration}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$. We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configurations that can be generated.

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
	\includegraphics[width=\textwidth]{figures/Fig2.png}
 \caption{\textbf{Four examples of grids produced by the spatial generator.} The lighter the red, the denser the area. Changing the growth rate $N$ allows to have more or less chaotic shapes (two first compared to the two last grids for example) corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$. Parameter values used here are for the first grid $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 971)$, the second $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 176)$, the third $(\alpha = 1.4,\beta=0.045, N = 102, n_d = 2, t_f = 618)$, and the fourth $(\alpha = 1.8,\beta=0.114, N = 108, n_d = 1, t_f = 227)$.}
\label{fig:spatialGen}
\end{figure} %
%%%%%%%%%%%%%%%%%%%%%%

In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure: their concentration index, hierarchy index, centrality index and homogeneity index (cf. \cite{LeNechet2015}). These morphological dimensions are in practice captured using the rank-size slope to capture hierarchy, the average distance between individuals to capture concentration, distribution entropy to capture homogeneity, and Moran spatial autocorrelation index to measure centrality.

We sample the generator parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a scatter with low discrepancy in high dimensions. We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}. It yields a subset of 170 grids matching empirical densities, which constituted our set of different initial spatial conditions. These are further clustered into three classes of morphology (figure \ref{fig:densityTypes}): compact (e.g. Vienna), polycentric (Liege) and discontinuous (Augsburg). This clustering allows to evaluate the non-trivial effects of a meaningful urban form on simulation results. We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model (cf. section \ref{sec:qualResults}).
